---
title: "BT weather data"
author: "Szymon Drobniak"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
      
      code_folding: hide
      toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE, cache = T}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, tidy = TRUE)
# options(digits = 2)
```

## Libraries

```{r libs, message = F}

## NOTE: loading warnings suppressed
## SET YOUR WD FIRST

library(pacman)
pacman::p_load(lme4, tidyverse,car)

library(here)
here()
```

## Processing of chick data

### Load data

```{r data_load}
data <- read.delim(here('data', 'EPC_WEATHER_2022.csv'), sep = ';', head = T)
temp <- read.table(here('data', '210805_temp.csv'), sep = ';', head = T)
rain <- read.table(here('data', '210805_rain.csv'), sep = ';', head = T)
data <- subset(data, !is.na(CODESTART) & !is.na(CODEEND))
```

### Process data

```{r processing, eval = F}

for (i in 1:nrow(data)) {
  
  # define the start timestamp
  codestart <- data[i, 'CODESTART']
  codeend <- data[i, 'CODEEND']
  
  cat(codestart, codeend, '\n')
  # if (codestart_prev == codestart)
  # cat(codestart, codeend, "\n")
  
  t_index1 <- which(temp[, "CODE"] == codestart)
  r_index1 <- which(rain[, "CODE"] == codestart)
  
  t_index2 <- which(temp[, "CODE"] == codeend)
  r_index2 <- which(rain[, "CODE"] == codeend)
  
  t_data <- temp[c(t_index1:t_index2),]
  r_data <- rain[c(r_index1:r_index2),]
  
  mean_temp <- mean(t_data[, "temp"], na.rm = T)
  sum_prcp <- sum(r_data[, "rain"], na.rm = T)
  
  data[i, "TEMP_BROOD"] <- mean_temp
  data[i, "PRCP_BROOD"] <- sum_prcp
  
  cat("Row "); cat(i); cat(" done\n")
}
```

```{r}
# write.table(data, row.names = F,
#             file = here::here("data", "postprocess", "BT_data_weather.csv"),
#             sep = ";")
```


## Processing of nest data

### Load data

```{r data_load_nests}
data <- read.delim(here('data', 'EPP WEATHER_nests.csv'), sep = ';', head = T)
temp <- read.table(here('data', '210805_temp.csv'), sep = ';', head = T)
rain <- read.table(here('data', '210805_rain.csv'), sep = ';', head = T)
data <- subset(data, !is.na(CODESTART) & !is.na(CODEEND))
```

### Process data for window t-14

Below code shows example processing of $t_{-14}$ time window, similar procedure can be used to obrain the $t_{-30}$ window. Do not run this if not needed (post-processed data is alrady in the foldser).

```{r processing_14days, eval = F}

for (i in 1:nrow(data)) {
  
  # define the start timestamp
  codestart <- data[i, 'CODESTART']
  codeend <- data[i, 'CODEEND']
  
  cat(codestart, codeend, '\n')
  # if (codestart_prev == codestart)
  # cat(codestart, codeend, "\n")
  
  t_index1 <- which(temp[, "CODE"] == codestart)
  r_index1 <- which(rain[, "CODE"] == codestart)
  
  t_index2 <- which(temp[, "CODE"] == codeend)
  r_index2 <- which(rain[, "CODE"] == codeend)
  
  t_data <- temp[c(t_index1:t_index2),]
  r_data <- rain[c(r_index1:r_index2),]
  
  mean_temp <- mean(t_data[, "temp"], na.rm = T)
  sum_prcp <- sum(r_data[, "rain"], na.rm = T)
  min_temp <- min(t_data[, "temp"], na.rm = T)
  max_temp <- max(t_data[, "temp"], na.rm = T)
  cv_temp <- sd(t_data[, "temp"]+273, na.rm = T)/mean(t_data[, "temp"]+273, na.rm = T)
  
  min_prcp <- min(r_data[, "rain"], na.rm = T)
  max_prcp <- max(r_data[, "rain"], na.rm = T)
  cv_prcp <- sd(r_data[, "rain"], na.rm = T)/mean(r_data[, "rain"], na.rm = T)
  
  
  data[i, "TEMP_BROOD14_mean"] <- mean_temp
  data[i, "PRCP_BROOD14_sum"] <- sum_prcp
  data[i, "TEMP_BROOD14_min"] <- min_temp
  data[i, "TEMP_BROOD14_max"] <- max_temp
  data[i, "TEMP_BROOD14_cv"] <- cv_temp
  data[i, "PRCP_BROOD14_min"] <- min_temp
  data[i, "PRCP_BROOD14_max"] <- max_temp
  data[i, "PRCP_BROOD14_cv"] <- cv_temp
  
  cat("Row "); cat(i); cat(" done\n")
}
```

```{r}
# write.table(data, row.names = F,
#             file = here::here("data", "postprocess", "BT_data_weather.csv"),
#             sep = ";")
```

## Modelling

Load post-processed data:

```{r}
dataset <- read_delim(here("data","postprocess","BT_data_weather_nests.csv"), ";",
                      escape_double = FALSE, trim_ws = TRUE)
# summary(dataset)
# glimpse(dataset)
```

Fit two main models with relevant variables:

```{r}
# model z 14-dniowym oknem przed LD
model1 <- glmer(EPC ~ scale(TEMP_BROOD14_mean)+scale(PRCP_BROOD14_sum)+
                  #I(TEMP_BROOD14_max - TEMP_BROOD14_min)+ #to byla amplituda temp ale usunalem ja
                  scale(TEMP_BROOD14_var) +
                  scale(PRCP_BROOD14_var) +
                  (1|FRING),
                family = 'binomial',
                data = dataset)

summary(model1)


# model z 30-dniowym oknem
model2 <- glmer(EPC ~ scale(TEMP_BROOD30_mean)+scale(PRCP_BROOD30_sum)+
                  #I(TEMP_BROOD14_max - TEMP_BROOD14_min)+ #to byla amplituda temp ale usunalem ja
                  scale(TEMP_BROOD30_var) +
                  scale(PRCP_BROOD30_var) +
                  (1|FRING),
                family = 'binomial',
                data = dataset)

summary(model2)
```

Plot predicted curves:

```{r}
plot(EPC ~ scale(TEMP_BROOD14_var), data = dataset)
curve(exp(-0.76+0.35*x)/(exp(-0.76+0.35*x)+1), col = 'red', lwd = 2, add = T)

plot(EPC ~ scale(TEMP_BROOD30_var), data = dataset)
curve(exp(-0.77+0.41*x)/(exp(-0.77+0.41*x)+1), col = 'red', lwd = 2, add = T)
```

## Explore variable correlations

Variance inflation factors:

```{r}
vif(model1)
vif(model2)
```

In all cases VIF<5, whcih means there are no collinearity problems.

Let's have a look at correlations across weather variables:

```{r}
pairs(dataset[,c(25,26,29,32)])
pairs(dataset[,c(33,34,37,40)])

cor(dataset[,c(25,26,29,32)])
cor(dataset[,c(33,34,37,40)])
```

The only worrying correlation exists between precipitation sum and variance (which is expected, sum of precipitation is similar to a Poisson variable).

Let's see if PCA could generate some meaningful scores:

```{r}
mypca_14 <- prcomp(dataset[,c(25,26,29,32)], center = T, scale. = T)
biplot(mypca_14)
biplot(mypca_14, choices = c(2,3))
```

Correlations between mean and variance clearly distinguish the thermal and precipitation dimentions, let's see what happens if we use PCs in the model:

```{r}
model3 <- glmer(EPC ~ mypca_14$x[,1] + mypca_14$x[,2] +
                  mypca_14$x[,3] + mypca_14$x[,4] +
                  (1|FRING),
                family = 'binomial',
                data = dataset)

summary(model3)
```

Results are the same: EPC is predicted by PC3 which is negatively loaded by temperature variability:

```{r}
mypca_14$rotation
```


```{r}
mypca_30 <- prcomp(dataset[,c(33,34,37,40)], center = T, scale. = T)
biplot(mypca_30)
biplot(mypca_30, choices = c(2,3))
```

```{r}
model4 <- glmer(EPC ~ mypca_30$x[,1] + mypca_30$x[,2] +
                  mypca_30$x[,3] + mypca_30$x[,4] +
                  (1|FRING),
                family = 'binomial',
                data = dataset)

summary(model4)
```

Conclusions are similar, this time PC3 is positively loaded by temperature variance and this is reflected in the GLMM output:

```{r}
mypca_30$rotation
```

## Interactions and simpler models

```{r}
# model z 14-dniowym oknem przed LD
model1.1 <- glmer(EPC ~ scale(TEMP_BROOD14_mean)+
                  scale(TEMP_BROOD14_var) +
                  (1|FRING),
                family = 'binomial',
                data = dataset)

summary(model1.1)


model1.2 <- glmer(EPC ~ scale(TEMP_BROOD14_var) +
                  (1|FRING),
                family = 'binomial',
                data = dataset)

summary(model1.2)

model1.3 <- glmer(EPC ~ scale(TEMP_BROOD14_mean)*scale(PRCP_BROOD14_sum)+
                  scale(TEMP_BROOD14_var)*scale(PRCP_BROOD14_var) +
                    scale(TEMP_BROOD14_mean)*scale(TEMP_BROOD14_var) +
                    scale(PRCP_BROOD14_var)*scale(PRCP_BROOD14_sum) +
                  (1|FRING),
                family = 'binomial',
                data = dataset)

summary(model1.3)

model1.3 <- glmer(EPC ~ scale(TEMP_BROOD14_mean) + scale(PRCP_BROOD14_sum)+
                  scale(TEMP_BROOD14_var)*scale(PRCP_BROOD14_var) +
                  (1|FRING),
                family = 'binomial',
                data = dataset)

summary(model1.3)
```

```{r}
# model z 14-dniowym oknem przed LD
model2.1 <- glmer(EPC ~ scale(TEMP_BROOD30_mean)+
                  scale(TEMP_BROOD30_var) +
                  (1|FRING),
                family = 'binomial',
                data = dataset)

summary(model2.1)


model2.2 <- glmer(EPC ~ scale(TEMP_BROOD30_var) +
                  (1|FRING),
                family = 'binomial',
                data = dataset)

summary(model2.2)

model2.3 <- glmer(EPC ~ scale(TEMP_BROOD30_mean)*scale(PRCP_BROOD30_sum)+
                  scale(TEMP_BROOD30_var)*scale(PRCP_BROOD30_var) +
                    scale(TEMP_BROOD30_mean)*scale(TEMP_BROOD30_var) +
                    scale(PRCP_BROOD30_var)*scale(PRCP_BROOD30_sum) +
                  (1|FRING),
                family = 'binomial',
                data = dataset)

summary(model2.3)

model2.3 <- glmer(EPC ~ scale(TEMP_BROOD30_mean) + scale(PRCP_BROOD30_sum)+
                  scale(TEMP_BROOD30_var)*scale(PRCP_BROOD30_var) +
                  (1|FRING),
                family = 'binomial',
                data = dataset)

summary(model2.3)
```

General conclusion: results are stable and there is no sign for any stron interaction.